---
title: "Incarceration in the United States"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    social: menu
    source_code: https://github.com/annmariebarrett/FinalProject-EDLD2/
    theme: readable
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(rio)
library(here)
library(janitor)
library(cowplot)
library(colorblindr)
#remotes::install_github("hrbrmstr/albersusa")
library(albersusa)
library(gganimate)
library(transformr)
library(ggthemes)
```

Data Source {.sidebar}
=================================
**Data Description**

These data depict the status and conditions of mass incarceration in the United States. Each plot is described below:

Plot 1:
This graph depicts the hourly wages offered in each state in 2017 for incarcerated people who are assigned to work in regular prison jobs and in state-owned businesses. The full brief can be found [here](https://www.prisonpolicy.org/blog/2017/04/10/wages/). 

Plot 2:
This line graph displays the rates of incarceration per 100,000 people over time within federal prisons, state prisons, and local jails. Data from local jails became available starting in 1983.  

Plot 3:
This animated map uses darkening colors to showcase the increasing rates of incarceration in state prisons over time in each state. 

**Data Source**

These visualizations use publically available data compiled by the [Prison Policy Initiative](https://www.prisonpolicy.org/data/). Plots on this page were created by [Ann-Marie Barrett](https://github.com/annmariebarrett) using the [ggplot](https://ggplot2.tidyverse.org/reference/ggplot.html) package in R.  

# Earnings

Column {.tabset .tabset-fade}
-----------------------------------------------------------------------

### Earnings By State

```{r, fig.height=8.5, fig.width=11}
#Import and clean earnings data
earnings <- import(here::here("data", "Earnings_state.xlsx")) %>% 
  characterize() %>% 
  clean_names() %>% 
  na_if('n/a') %>% 
  mutate(non_industry = as.double(non_industry), 
         correctional_industry = as.double(correctional_industry))

#Plot earnings final version
p1 <- ggplot(earnings, 
             aes(y = state, 
                 x = non_industry)) +
  geom_line(aes(group = state)) +
  geom_point(aes(color = bookend), 
             show.legend = FALSE) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(title = "...and in Non-Industry Jobs", 
       x = "", 
       y = "") + 
  theme_solarized_2() +
  theme(text = element_text(color = "grey30"), 
        title = element_text(color = "grey30"),
        plot.margin = margin(.5, 1.5, .5, .5, unit = "cm")) +
  scale_x_continuous(limits = c(0, 5.15), labels=scales::label_dollar()) +
  scale_y_discrete(limits = rev)
  
p2 <- ggplot(earnings, 
             aes(y = state, 
                 x = correctional_industry)) +
  geom_line(aes(group = state)) +
  geom_point(aes(color = bookend), 
             show.legend = FALSE) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(title = "Hourly Wage in Correctional Industry Jobs", 
       x = "",
       y = "State") +       
  theme_solarized_2(light = T) +
  theme(text = element_text(color = "grey30"), 
        title = element_text(color = "grey30"), 
        axis.title.y = element_text(size = 16),
        plot.margin = margin(.5, .5, .5, .5, unit = "cm")) +
  scale_x_continuous(labels=scales::label_dollar()) + 
  scale_y_discrete(limits = rev)

plot_grid(p2, p1, rel_widths = c(1, 1.1))

#Possible improvements: Combining the plots such that you see the earnings for both groups on one plot. So each state would have 2 rows. I think incorporating the color green could be really nice since your outcome is money #Moolah and coordinating so that colors in the title match the colors used to indicate the group type.  
```

### Version 2

```{r fig.height=8.5, fig.width=11}
#Plot earnings fig version 2
p5 <- ggplot(earnings, 
             aes(y = state, 
                 x = non_industry)) +
  geom_line(aes(group = state)) +
  geom_point(aes(color = bookend), 
             show.legend = FALSE) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(title = "...and in Non-Industry Jobs", 
       x = "Wage per  hour", 
       y = "") + 
  theme_minimal() +
  scale_x_continuous(limits = c(0, 5.15), labels=scales::label_dollar()) 

p6 <- ggplot(earnings, 
             aes(y = state, 
                 x = correctional_industry)) +
  geom_line(aes(group = state)) +
  geom_point(aes(color = bookend), 
             show.legend = FALSE) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(title = "Earnings in Correctional Industry Jobs", 
       x = "Wage per hour",
       y = "State") +       
  theme_minimal() +
  scale_x_continuous(labels=scales::label_dollar())

plot_grid(p6, p5, rel_widths = c(1, 1.1))
```

### Version 1

```{r, fig.height=6.5, fig.width=14}
#Plot earnings fig version 1
p3 <- ggplot(earnings, 
             aes(y = state, 
                 x = non_industry)) +
  geom_line(aes(group = state)) +
  geom_point(aes(color = bookend)) +
  xlim(0, 5.15) +
  labs(title = "Range of Earnings Offered to Incarcerated People in Non-industry Jobs", 
       x = "Wage per  hour", 
       y = "State", 
       caption = "Data from Sawyer (2017). https://www.prisonpolicy.org/blog/2017/04/10/wages/"
       ) 

p4 <- ggplot(earnings, 
             aes(y = state, 
                 x = correctional_industry)) +
  geom_line(aes(group = state)) +
  geom_point(aes(color = bookend)) +
  labs(title = "Range of Earnings Offered to Incarcerated People in Correctional Industry Jobs", 
       x = "Wage per hour", 
       y = "State", 
       caption = "Data from Sawyer (2017). https://www.prisonpolicy.org/blog/2017/04/10/wages/") 

plot_grid(p3, p4)
```

Incarceration Rates
===================================

Column {.tabset .tabset-fade}
-----------------------------------------------------------------------

### Rates Over Time 

```{r, fig.height = 6, fig.width=8}
#import incarceration rate data
incar <- import(here::here("data", "incarceration_rates_by_type.csv")) %>% 
  characterize() %>% 
  clean_names()

#get rid of empty cells  
incar <- incar[-c(93:16383), ]

#plot animated line graph 
p5 <- incar %>%
  pivot_longer(cols = state_prisons:local_jails) %>%
  ggplot(aes(year, value, color = name)) +
  geom_line(size = 1.25) +
  labs(y = "Incarceration Rates",
       x = "",
       title = "Rates of Incarceration per 100,000 people \nin the United States",
       subtitle = "1925 - 2016") +
  scale_color_OkabeIto(name = "Institution",
                       labels = c("Federal Prisons",
                                  "Local Jails",
                                  "State Prisons")) +
  theme_solarized_2() +
  theme(text = element_text(color = "grey30"),
        title = element_text(color = "grey30"),
        plot.margin = margin(.5, .5, .5, .7, unit = "cm"),
        axis.title.y = element_text(margin = margin(0, 15, 0, 0))) +
  transition_reveal(year)

animate(p5, renderer = magick_renderer())

#Code to save .gif
#anim_save("rates_linegraph.gif", p5, renderer = magick_renderer())
```

### Rates Over Time: Option B

```{r fig.height = 6, fig.width=7}
#rearrange data to more easily feed into this style of line plot
incar_2 <- incar %>% 
  pivot_longer(cols = state_prisons:local_jails) %>% 
  mutate(Type = recode(name, "federal_prisons" = "Federal Prisons", "state_prisons" = "State Prisons", "local_jails" = "Local Jails"))

#plot static line graph
incar_2 %>% 
  ggplot(aes(year, value, color = Type)) +
  geom_line(size = 1.25) +
  labs(y = "Incarceration Rates", 
       x = "", 
       title = "Rates of Incarceration per 100,000 people \nin the United States",
       subtitle = "1925 - 2016") +
  scale_color_OkabeIto(name = "Institution", 
                       labels = c("Federal Prisons", 
                                  "Local Jails", 
                                  "State Prisons")) + 
  theme_solarized_2() +
  geom_text(data = filter(incar_2, year == 2016), 
            aes(y = value, label = Type), 
            nudge_x = 2, 
            hjust = 0, 
            color = "grey30") +
  guides(color = "none") +
  xlim(1925, 2034) +
  theme(text = element_text(color = "grey30"), 
        title = element_text(color = "grey30"),
        plot.margin = margin(.5, .5, .5, .7, unit = "cm"),
        axis.title.y = element_text(margin = margin(0, 15, 0, 0)),
        panel.background = element_rect(fill = NA),
        panel.grid.major = element_line(color = "grey90"))
```

### Version 2

```{r fig.height = 6, fig.width=7}
#second version of static line graph
incar %>% 
  pivot_longer(cols = state_prisons:local_jails) %>%
  ggplot(aes(year, value, color = name)) +
  geom_line() +
  labs(y = "Incarceration Rates", 
       x = "Year", 
       title = "Rates of Incarceration per 100,000 people", 
       caption = "Data compiled by the Prison Policy Initiative. https://www.prisonpolicy.org/data/") +
  scale_color_OkabeIto(name = "Institution", 
                       labels = c("Federal Prisons", 
                                  "Local Jails", 
                                  "State Prisons")) + 
  theme_minimal(base_size = 18) 
```

### Version 1

```{r}
#first version of static line graph
incar %>% 
  pivot_longer(cols = state_prisons:local_jails) %>%
  ggplot(aes(year, value, color = name)) +
  geom_line() +
  labs(y = "Incarceration Rates", 
       x = "Year", 
       title = "Rates of Incarceration per 100,000 people", 
       caption = "Data compiled by the Prison Policy Initiative. https://www.prisonpolicy.org/data/")
```


Map of Increasing Rates
===================================

Column {.tabset .tabset-fade}
-----------------------  

### Map of Rates Over Time

```{r}
#Import Data
states <- import(here::here("data", "State_rates_time.xlsx")) %>% 
  characterize() %>% 
  clean_names()

#Join map and state prison rate data
us <- usa_sf() %>% 
  mutate(
    state = factor(name))

states_prison <- states %>% 
  filter(institution == "Prison Rate")

states_prison[3:40] <- sapply(states_prison[3:40], as.numeric)

states_prison <- states_prison %>% 
  pivot_longer(cols = starts_with("x"), names_to = "year", values_to = "rates")

states_prison$year <- gsub("[^0-9.-]", "", states_prison$year)

states_prison <- states_prison %>% 
  mutate(year = factor(year))

data_map <- left_join(us, states_prison, by = "state")

#Plot animated map
plot_map <-
  ggplot(data_map) +
  geom_sf(aes(fill = rates)) +
  scale_fill_distiller(palette = "YlOrBr", direction = 1) +
  labs(title = "Incarceration in State Prisons in {closest_state}",
       fill = "Rates per 100,000 people") +
  theme_solarized_2() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        text = element_text(color = "grey30"),
        title = element_text(color = "grey30"),
        panel.background = element_blank()) +
  theme(legend.position = 'bottom') +
  transition_states(year,
                    transition_length = 3,
                    state_length = 3) +
  ease_aes('cubic-in-out')

animate(plot_map, renderer = magick_renderer())

#Code to save .gif
# anim_save("map.gif", plot_map, renderer = magick_renderer())
```

### Version 2

```{r}
#Second version of animated map 
plot_map2 <-
  ggplot(data_map) +
  geom_sf(aes(fill = rates)) +
  scale_fill_viridis_c(option = "plasma") +
  labs(title = "Rates of Incarceration in State Prisons  in {closest_state}",
       fill = "Incarceration per 100,000 people",
       caption = "Data from Sawyer (2018). https://www.prisonpolicy.org/reports/women_overtime_table_1.html") +
  theme_minimal() +
  theme(legend.position = 'top',
        text = element_text(color = "grey30"),
        title = element_text(color = "grey30")) +
  transition_states(year,
                    transition_length = 3,
                    state_length = 3) +
  ease_aes('cubic-in-out')

animate(plot_map2, renderer = magick_renderer())
```

### Version 1

```{r}
#First version - static map
  ggplot(data_map) +
  geom_sf(aes(fill = rates)) +
  labs(title = "Rates of Incarceration in State Prisons", 
       fill = "Incarceration per 100,000 people",
       caption = "Data from Sawyer (2018). https://www.prisonpolicy.org/reports/women_overtime_table_1.html")
```

